Interactive Graphics for Visually Diagnosing Forest Classifiers in R

### 2022-06-28 PPforestVizupdate.Rnw
Fix some typos

\begin{abstract}
This article describes structuring data and constructing plots to explore forest classification models interactively. A forest classifier is an example of an ensemble since it is produced by bagging multiple trees. The process of bagging and combining results from multiple trees produces numerous diagnostics which, with interactive graphics, can provide a lot of insight into class structure in high dimensions. Various aspects are explored in this article, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations. The ideas are applied to the random forest algorithm and projection pursuit forest but could be more broadly applied to other bagged ensembles helping in the interpretability deficit of these methods. Interactive graphics are built in R using the \texttt{ggplot2}, \texttt{plotly}, and \texttt{shiny} packages.
\keywords{Statistical Visualization \and Interactive Visualization \and Interpretable Machine Learning \and Ensemble Model}
\end{abstract}

@@

Ensemble classifiers have grown in popularity~\citep{dietterish00}~\citep{talbot09}, and the basic ideas behind the random forest can be applied to virtually any type of model. The benefits for classification are reduced variability in predictive error, and the suite of diagnostics provides the potential for better understanding the class structure in the high-dimensional data space. The use of visualization on these diagnostics, in association with multivariate data plots, completes the process to better understand the underlying problem.

@@

There has been some, but not a lot of, research on visualizing classification models.
\cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \verb# KLIMT# that includes zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps. \cite{cutler15raft} developed a java package called \verb# RAFT# to visualize a forest classifier, that included variable selection, parallel coordinate plots, heat maps, and scatter plots of some diagnostics. Linking between plots is limited. \cite{quach2012interactive} presents interactive forest visualization using the R package \verb# iPlots eXtreme # \citep{urbanek2011iplots} where several displays are shown in one window with some linking between them available. \cite{Silva2016} describes visualizing components of an ensemble classifier.

@@

The visualization approach is consistent with the framework in \cite{wickham2015visualizing}, and the implementation is built on the newest interactive graphics available in R \citep{RCore}. The purpose is to provide readily available tools for users to explore and improve ensemble fits and obtain an intuition for the underlying class structure in data. Interactive plots are a key component for model visualization that helps the user see multivariate relationships and be more efficient in the model diagnosis. Multiple levels of data are constructed for exploration: observation, model, and ensemble summaries.
The visualization tools presented in this article help open up black-box models and contribute to the interpretability of these methods.

@@

@
 \vspace*{-0.3cm}
 \caption{Decision boundaries for PPtree algorithm on 2D simulated data. The partitions generated by PPtree algorithm are oblique to the axis, incorporating the association between the two variables.\label{bounds}}
\end{figure}
Two data set are used in this paper, crab ~\citep{CM74} and fishcatch data~\citep{fishcatch}. To illustrate the mapping of diagnostics to visualization crab data are used because its simplicity. The second data set is introduced to illustrate the interactive web app which combines the visualizations introduced in this paper and present a more complex problem to explore within the applicarion.

The article is structured as follows.  Section \ref{key} describes the ensemble components to be accessed.  Section \ref{visen} maps the ensemble components to the visual elements. The web app is described in \ref{app} and further work is discussed in Section \ref{fur}.

@@

\item {\em \textbf{Uncertainty measure for each observation:}} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class. If a case is always predicted to be the true class, there is no uncertainty about an observation. Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations. They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or more simply, they may be errors in measurements in the data.\\

@@

\item {\em \textbf{Similarity measure for pairs of observations:}} In each model, each pair of observations will be either in the same terminal node or not. This is used to compute a proximity matrix.  Cluster analysis on this matrix can be used to follow up the classification to assess the original labeling. It may suggest improvements or mistakes in original labels.
\end{itemize}
In addition to these overall ensemble statistics, each component model has its own diagnostics, measuring error, variables utilized, and class predictions. Visualization will enable the individual models to be examined, relate these to the data and their contribution to the ensemble.

@@

The PPF is composed of individual projection pursuit trees, PPtree algorithm uses a multi-step approach to fit a multi-class model by finding linear combinations to split on based on projection pursuit algorithm. Two projection pursuit indexes are used to find projections that separate classes,  LDA  \citep{lee2005projection} and PDA \citep{lee2010projection}.

@@

\noindent where notation is the same as for the LDA index, with the addition of $\lambda \in [0,1)$ is a shrinkage parameter, and a different within group sum of squares, $\blW_{PDA}(\lambda)=\mbox{diag}(\blW)+(1-\lambda)\mbox{offdiag}(\blW)$.

Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data using LDA index (Equation \ref{lda}) and two variables in each node split. There are three nodes for the four class problem. The nodes of this tree are based on projections of the data; the coefficients of which form the building block to calculate the variable importance. The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
The package \verb#PPtreeViz# provides visual tools to diagnose a PPtree model. The PPF builds on these and modifies a little. The PPtree model is simpler than a regular classification tree because the classes are mostly separated by combinations of variables -- just three projections are needed to see the differences between the four classes.

@@

@
\caption{Visualizing the PPtree model of the crab data. The tree has three nodes (top). The density plots show the data projections at each node, colored by group (middle). The dashed vertical red line indicates the split value of each node. At node 1 separates the sexes. Nodes 2 separate males from species and node 3 separate females from species. Mosaic plots of the confusion table for each split (bottom). Node 1 shows the clear split of sexes, with a small number of misclassifications. Node 2 where blue males are separated from orange males with small number of misclassifications. Node 3 where blue females are separated from orange females, almost perfect classification.\label{pptreeviz1}}
\end{figure}

@@

\noindent where $\hat y_{ik}$ is the predicted class for the observation $i$ in the tree $k$, and $\tilde y_{ik}$ is the predicted class for the observation $i$ in the tree $k$ after permuting the values for variable $j$. The global permuted importance measure is the average importance over all the trees in the forest. This measure is based on comparing the accuracy of classifying oob observations using the true class with permuted (nonsense) class.

@@

\noindent where $e_k$ is the oob error rate of tree $k$, $I_{sjk}$ is the projection pursuit index value of node $s$ ~\cite{da2021projection}.

Figure \ref{impotree} shows the absolute projection coefficient of the top three nodes for all the trees in a forest model. This information is displayed by a side-by-side jittered dot plot. The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptreeviz1}. The forest was built using random samples of two variables for each node, hence there are two coefficients for each node. At node 1 uses CW and RW with similar contribution. The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce. Node 2 uses BD and CL, BD contributes the most to the separation. The plot in the right shows the boundary that is induced. Node 3 uses CW and FL, and with an even contribution by the two variables.
For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.

@@

@
\caption{Visualizing the variable importance of all the trees in the forest, for three nodes. Each node of each tree has an importance value for each variable. The values for the whole forest are displayed using a side-by-side jittered dot plot. The importance values are the absolute values of the projection coefficients. The red points correspond to these values for the tree shown in Figure \ref{pptreeviz1}. Two variables are randomly selected at each node for creating the best projection, and split. The plots at right show the variables used and the split made at each of the nodes of this tree.  \label{impotree}}
\end{figure}

@@

For each tree, every pair of observations can be in the same terminal node or not. Tallying this up across all trees in a forest gives the proximity matrix, an $n\times n$ matrix of the proportion of trees that the pair share a terminal node. It can be considered a similarity matrix.

Multidimensional scaling (MDS) is used to reduce the dimension of this matrix, to view the similarity between observations ~\citep{kruskal1964multidimensional}. MDS transforms the dataset into a low-dimensional space where the distances are approximately the same as in the full $n$ dimensions. With $G$ groups, the low-dimensional space should be no more than $G-1$ dimensions. Figure \ref{prox1} shows the MDS plots for the three dimensional space induced by the four groups of the crab data. Color indicates the true species and sex. For this data two dimensions are enough to see the four groups separated quite well. Some crabs are clearly more similar to a different group, especially in examining the sex differences. MDS is included in the interactive web app presented in Section \ref{app}.

@@

@
\caption{Generalized ternary plot ((G-1) dimensional simplex, here it is a tetrahedron) representation of the vote matrix for four classes. The tetrahedron is shown pairwise. Each point corresponds to one observation and color is the true class. This is close but not a perfect classification since the colors are concentrated in the corners and there are some mixed colors in each corner.}
\label{tetra}
\end{figure}

Because visualizing the vote matrix with a $(G-1)$ dimensional tetrahedron requires dynamic graphics, a low-dimensional option is also provided. For each class, each case has a value between 0-1. A side-by-side jittered dotplot is used for the display, where the class is displayed on one axis and proportion is displayed on the other. For each dotplot, the ideal arrangement is that points are concentrated at 0 or 1 and only at 1 for their true class. This data is close to the ideal but not perfect, e.g. there are a few blue male crabs (orange) that are frequently predicted to be blue females (green), and a few blue female crabs predicted to be another class.

@@

 Interaction is added to the plots described in Section \ref{visen} and other plots, and they are organized into an interactive web app using \verb#shiny#~\citep{chang11shiny, wickham2021mastering} for exploring the ensemble model. The app is organized into three tabs: individual cases; models; and performance comparison, to provide a model diagnostic tool. Interaction is provided as mouse-over labeling, mouse-click selection, and brushing, with results linked across multiple plots. The app takes advantage of new tools provided in the \verb#plotly#~\citep{plotly} package, developed as a part of Sievert's PhD thesis research~\citep{sievertthesis}.

@@

To illustrate the shiny app characteristics a different data (fishcatch) are used which present a more challenging classification problem and then more interesting results to explore the app. The fishcatch data~\citep{fishcatch} contains 159 observations, with 6 physical measurement variables, and  7 types of fish, all caught from the same lake (Laengelmavesi) near Tampere in Finland. There are 35 bream, 11 parkki, 56 perch, 17 pike, 20 roach, 14 smelt and 6 whitewish. The shiny app showing fishcatch data can be accessed at \url{https://natydasilva.shinyapps.io/shinyppforest}.

@@

This selection of plots enables aspects of the model, relating to performance for individual cases, to be examined in the data space. The data plot is an essential element following the {\em model-in-the-data-space} philosophy of \citet{wickham2015visualizing}. The choice was made to use a parallel coordinate plot because it provides a space-efficient display of the data. Alternatives include the tour, a dynamic plot, or a scatterplot matrix. Theoretically, either of these could be substituted or added. 
Parallel coordinate plots are useful to visualize multivariate numerical data. As is shown in the first plot in Figure \ref{tab1}, in a PCP each variable is placed equidistant and perpendicular to the x-axis, on the y-axis the values for each variable are standardized to easily compare them. Each observation is represented as a line connected across  all the variables. In a PCP many variables can be compared together and seeing the relationships between them, identify and characterize groups of observations. The features can be ordered so that there are not too many lines intersecting resulting in an unreadable chart \citet{inselberg2009parallel}. 
Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced. Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}. Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variable pairs are shown next to each other once.

@@

Figure \ref{tab2} is a screenshot of the models tab. There are five plots, with varying levels of interaction: (1) a boxplot of oob error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.  The interaction is driven in three ways, from the boxplot -- when the user selects a point in that display, the corresponding importance variable plot, tree plot from the \verb#PPtreeViz#, tree density displays and mosaic plots are drawn. The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure. Also the variable importance values are highlighted for each variable for each of the top three nodes, by default, and the oob error value for the selected tree on the boxplot (both in red).
Additionally it is possible to select specific nodes and update the importance variable plot, density plot and mosaic plot (nodes can be selected on the right corner on this second tab) and finally the interaction can be driven from the variable importance plot.

The selected tree in Figure \ref{tab2} has very small oob error shown as red point in the boxplot. Two variables were use in each node partition, the first four nodes were selected and display in the importance variable plot, density plot and mosaic plot. In the importance variable plot `height` and `lenght1` were used in the first node partition being `height` the important variable to separate Bream and Parkki (Green and orange in density and mosaic plots) from the rest. In the second node `length1` and `length2` were both evenly important  to separate Bream from Parkki. A similar analysis can be done for other nodes.

@@

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/fishT2.pdf}
\caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app. The boxplot and the dotplot of variable importance has click selection, which invokes changes of the corresponding displays. Selecting a point in the boxplot update the node tree information (Shiny Gui) and update all the plots for the three first nodes, also nodes can be selected and update the corresponding plots. Finally selecting a point in the importance variable plot makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
\label{tab2diag}
\end{figure}

\subsection{Performance Comparison}
The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit. There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) a receiver operating characteristic curve (ROC) curve comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance. There is very little interaction on this tab. Users can select to focus on a subset of classes or choose the importance measure to show. Being able to focus on class can help to better understand how well the model performs across classes and the focus can be especially useful for unbalanced data. Examining the oob error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.

@@

Having better tools to open up black box models will provide for better understanding the data, the model strengths and weaknesses, and how the model will perform on future data. This visualisation app provides a selection of interactive plots to diagnose PPF models. This shell could be used to make an app for other ensemble classifiers. The philosophy underlying the collection of displays is ``show the model in the data space''  explained in \citet{wickham2015visualizing}. It is not easy to do this, and to completely take this on would require plotting the model in the $p$-dimensional data space. In the simplest approach, as taken here, it means to link the model diagnostics to displays of the data. Then it is possible to probe and query to obtain a better understanding such as finding regions in the data that prove difficult to fit, and detract from the predictive accuracy, or that don't adhere to model assumptions.

@@

The application have some limitations that are need to pointed out, if the number of observation is big the overploting will be an issues in some of the selectedd visualizations and if the number of variables increases, plots like parallel plot and importance variable plot should include the option to select some variables each time or a different approach.

There are many possible extensions to the app, that could help it to be a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem. Working within the R environment makes all of these desires available using command line outside the app, given the unique IDs of models and cases can be exported from the app.

The app has helped to identify ways to improve the  PPtree algorithm and consequently, the PPF model. These especially apply to multiclass problems. Multiple splits for the same class would enable nonlinear classifications. Split criteria tend to place boundaries too close to some groups, due to heteroskedasticity being induced by aggregating classes. Forests are not always better than their constituent trees, and if the trees can be built better, the forest will provide stronger predictions.

>>>>

\begin{abstract}
This article describes structuring data and constructing plots to explore forest classification models interactively. A forest classifier is an example of an ensemble since it is produced by bagging multiple trees. The process of bagging and combining results from multiple trees produces numerous diagnostics which, with interactive graphics, can provide a lot of insight into class structure in high dimensions. Various aspects of models are explored in this article, to assess model complexity, individual model contributions, variable importance and dimension reduction, and uncertainty in prediction associated with individual observations. The ideas are applied to the random forest algorithm and projection pursuit forest but could be more broadly applied to other bagged ensembles helping in the interpretability deficit of these methods. Interactive graphics are built in R using the \texttt{ggplot2}, \texttt{plotly}, and \texttt{shiny} packages.
\keywords{Statistical Visualization \and Interactive Visualization \and Interpretable Machine Learning \and Ensemble Model}
\end{abstract}

@@

Ensemble classifiers have grown in popularity~\citep{dietterish00, talbot09}, and the basic ideas behind the random forest can be applied to virtually any type of model. The benefits for classification are reduced variability in predictive error, and the suite of diagnostics provides the potential for better understanding the class structure in the high-dimensional data space. The use of visualization on these diagnostics, in association with multivariate data plots, completes the process to better understand the underlying problem.

@@

There has been some, but not a abundant, research on visualizing classification models.
\cite{urbanek2002exploring} presents interactive tree visualization implemented in the java software called \verb# KLIMT# that includes zooming, selection, multiple views, interactive pruning, and tree construction as well as the interactive analysis of forests of trees using treemaps. \cite{cutler15raft} developed a java package called \verb# RAFT# to visualize a forest classifier, that included variable selection, parallel coordinate plots, heat maps, and scatter plots of some diagnostics. Linking between plots is limited. \cite{quach2012interactive} presents interactive forest visualization using the R package \verb# iPlots eXtreme # \citep{urbanek2011iplots} where several displays are shown in one window with some linking between them available. \cite{Silva2016} describes visualizing components of an ensemble classifier.

@@

The visualization approach is consistent with the framework in \cite{wickham2015visualizing}, and the implementation is built on the newest interactive graphics available in R \citep{RCore}. The purpose is to provide readily available tools for users to explore and improve ensemble fits and obtain an intuition for the underlying class structure in data. Interactive plots are a key component for model visualization that helps the user see multivariate relationships and be more efficient in model diagnosis. Multiple levels of data are constructed for exploration: observation, model, and ensemble summaries.
The visualization tools presented in this article help open up black-box models and contribute to the interpretability of these methods.

@@

@
 \vspace*{-0.3cm}
 \caption{Decision boundaries for PPtree algorithm on 2D simulated data. The partitions generated by PPtree algorithm are oblique to the axis, incorporating the association between the two variables.\label{bounds}}
\end{figure}
Two data set are used in this paper, crab ~\citep{CM74} and fishcatch data~\citep{fishcatch}. To illustrate the mapping of diagnostics to visualization crab data are used because of its simplicity. The second data set is introduced to illustrate the interactive web app which combines the visualizations introduced in this paper and present a more complex problem to explore within the application.

The article is structured as follows.  Section \ref{key} describes the ensemble components to be assessed.  Section \ref{visen} maps the ensemble components to the visual elements. The web app is described in Section \ref{app} and further work is discussed in Section \ref{fur}.

@@

\item {\em \textbf{Uncertainty measure for each observation:}} Across individual (classification) models we can compute the proportion of times that a case is predicted to be each class. If a case is always predicted to be the true class, there is no uncertainty about an observation. Cases that are predicted to be multiple classes, with similar class probabilities, indicate difficulty to classify observations. They may be important by indicating neighborhoods of the data space that would benefit from a more complex model, or simply, measurement errors in the data.\\

@@

\item {\em \textbf{Similarity measure for pairs of observations:}} In each model, a pair of observations will be either in the same terminal node or not. This is used to compute a proximity matrix. Cluster analysis on this matrix can be used to follow up the classification to assess the original labeling. It may suggest improvements or mistakes in original labels.
\end{itemize}
In addition to these overall ensemble statistics, each individual model has its own diagnostics, measuring error, variables utilized, and class predictions. Visualization will enable the individual models to be examined, relate these to the data and their contribution to the ensemble.

@@

The PPF is composed of individual projection pursuit trees. PPtree algorithm uses a multi-step approach to fit a multi-class model by finding linear combinations to split on based on projection pursuit algorithm. Two projection pursuit indexes are used to find projections that separate classes,  LDA  \citep{lee2005projection} and PDA \citep{lee2010projection}.

@@

\noindent where the notation is analogous to the LDA index, with the addition of $\lambda \in [0,1)$ which is a shrinkage parameter, and a different within group sum of squares, $\blW_{PDA}(\lambda)=\mbox{diag}(\blW)+(1-\lambda)\mbox{offdiag}(\blW)$.

Figure \ref{pptreeviz1} shows a visual ensemble of plots of a tree model on the crab data using LDA index (Equation \ref{lda}) and two variables in each node split. There are three nodes for the four class problem. The nodes of this tree are based on projections of the data and their coefficients form the building block to calculate the variable importance. The density plot displays the data projection at each node, and the mosaic plot shows the confusion matrix for the nodes.
The package \verb#PPtreeViz# provides visual tools to diagnose a PPtree model. The PPF builds on these and modifies a little. The PPtree model is simpler than a regular classification tree because the classes are mostly separated by combinations of variables -- just three projections are needed to see the differences between the four classes.

@@

@
\caption{Visualizing the PPtree model of the crab data. The tree has three nodes (top). The density plots show the data projections at each node, colored by group (middle). The dashed vertical red line indicates the split value of each node. Node 1 separates the sexes. Node 2 separate males from species and node 3 separate females from species. Mosaic plots of the confusion table for each split (bottom). Node 1 shows the clear split of sexes, with a small number of misclassifications. Node 2 where blue males are separated from orange males with small number of misclassifications. Node 3 where blue females are separated from orange females, almost perfect classification.\label{pptreeviz1}}
\end{figure}

@@

\noindent where $\hat y_{ik}$ is the predicted class for the observation $i$ in the tree $k$, and $\tilde y_{ik}$ is the predicted class for the observation $i$ in the tree $k$ after permuting the values for variable $j$. The global permuted importance measure is the average importance over all the trees in the forest. This measure is based on comparing the accuracy of classifying oob observations using the true class with permuted (nonsense) classes.

@@

\noindent where $e_k$ is the oob error rate of tree $k$, $I_{sjk}$ is the projection pursuit index value of node $s$ ~\citep{da2021projection}.

Figure \ref{impotree} shows the absolute projection coefficient of the top three nodes for all the trees in a forest model. This information is displayed by a side-by-side jittered dot plot. The red dots correspond to the absolute coefficient values for the tree model of Figure \ref{pptreeviz1}. The forest was built using random samples of two variables for each node, hence there are two coefficients for each node. At node 1 uses CW and RW with similar contribution. The scatterplot at right shows these two variables and the resulting boundary between groups that this would produce. Node 2 uses BD and CL where BD contributes the most to the separation. The plot in the right shows the boundary that is induced. Node 3 uses CW and FL with an even contribution by the two variables.
For each tree in the forest, different decision rules are defined; the resulting boundaries on the previous plots are based on Rule 1 $= \frac{m_1}{2} + \frac{m_2}{2} $, where $m_1$ and $m_2$ are the mean of the left and right groups at each node.

@@

@
\caption{Visualizing the variable importance of all the trees in the forest, for three nodes. Every particular node of each tree has an importance value for the corresponding variable. The values for the whole forest are displayed using a side-by-side jittered dot plot. The importance values are the absolute values of the projection coefficients. The red points correspond to these values for the tree shown in Figure \ref{pptreeviz1}. Two variables are randomly selected at each node for creating the best projection, and split. The plots at right show the variables used and the split made at each of the nodes of this tree.  \label{impotree}}
\end{figure}

@@

For each tree, every pair of observations can be in the same terminal node or not. Tallying this up across all trees in a forest gives the proximity matrix, an $n\times n$ matrix of the proportion of trees that the pair share a terminal node. It can be considered as a similarity matrix.

Multidimensional scaling (MDS) is used to reduce the dimension of this matrix, to view the similarity between observations ~\citep{kruskal1964multidimensional}. MDS transforms the dataset into a low-dimensional space where the distances are approximately the same as in the full $n$ dimensions. With $G$ groups, the low-dimensional space should be no more than $G-1$ dimensions. Figure \ref{prox1} shows the MDS plots for the three dimensional space induced by the four groups of the crab data. Color indicates the true species and sex. For this data, two dimensions are enough to see the four groups separated quite well. Some crabs are clearly more similar to a different group, especially in examining the sex differences. MDS is included in the interactive web app presented in Section \ref{app}.

@@

@
\caption{Generalized ternary plot ((G-1) dimensional simplex, in this case a tetrahedron) representation of the vote matrix for four classes. The tetrahedron is shown pairwise. Each point corresponds to one observation and color is the true class. This is close but not a perfect classification since the colors are concentrated in the corners and there are some mixed colors in each corner.}
\label{tetra}
\end{figure}

Because visualizing the vote matrix with a $(G-1)$ dimensional tetrahedron requires dynamic graphics, a low-dimensional option is also provided. For each class, every individual case has a value between 0-1. A side-by-side jittered dotplot is used for the display, where the class is displayed on one axis and proportion is displayed on the other. For each dotplot, the ideal arrangement is that points are concentrated at 0 or 1 and only at 1 for their true class. This data is close to the ideal but not perfect, e.g. there are a few blue male crabs (orange) that are frequently predicted to be blue females (green), and a few blue female crabs predicted to be another class.

@@

 Interaction is added to the plots described in Section \ref{visen} as well as other plots, which are organized into an interactive web app using \verb#shiny#~\citep{chang11shiny, wickham2021mastering} for exploring the ensemble model. The app is organized into three tabs: individual cases; models; and performance comparison, to provide a model diagnostic tool. Interaction is provided as mouse-over labeling, mouse-click selection and brushing, with results linked across multiple plots. The app takes advantage of new tools provided in the \verb#plotly#~\citep{plotly} package, developed as a part of Sievert's PhD thesis research~\citep{sievertthesis}.

@@

To illustrate the shiny app characteristics, a different dataset (fishcatch) are used which presents a more challenging classification problem and then more interesting results to explore the app. The fishcatch dataset~\citep{fishcatch} contains 159 observations, with 6 physical measurement variables, and  7 types of fish, all caught from the same lake (Laengelmavesi) near Tampere in Finland. There are 35 bream, 11 parkki, 56 perch, 17 pike, 20 roach, 14 smelt and 6 whitewish. The shiny app showing fishcatch data can be accessed at \url{https://natydasilva.shinyapps.io/shinyppforest}.

@@

This selection of plots enables aspects of the model, relating to performance for individual cases, to be examined in the data space. The data plot is an essential element following the {\em model-in-the-data-space} philosophy of \citet{wickham2015visualizing}. The choice was made to use a parallel coordinate plot because it provides a space-efficient display of the data. Alternatives include the tour, a dynamic plot, or a scatterplot matrix. Theoretically, either of these could be substituted or added. 
Parallel coordinate plots are useful to visualize multivariate numerical data. As is shown in the first plot in Figure \ref{tab1}, in a PCP each variable is placed equidistant and perpendicular to the x-axis, on the y-axis the values for each variable are standardized to easily compare them. Each observation is represented as a line connected across  all the variables. In a PCP many variables can be compared together and seeing the relationships between them, identify and characterize groups of observations. The features can be ordered so that there are not too many lines intersecting resulting in an unreadable chart \citep{inselberg2009parallel}. 
Two alternatives can be selected in shiny to draw the parallel coordinate plot: parallel or enhanced. Parallel draws the classic PCP and enhanced draws a modified version where variables are repeated \citep{hurley2011eulerian}. Because reading a PCP is really only possible for neighboring variables, the variables are repeated so that all variable pairs are shown next to each other once.

@@

Figure \ref{tab2} is a screenshot of the models tab. There are five plots, with varying levels of interaction: (1) a boxplot of oob error for all trees including the data as jittered points, (2) a jittered side-by-side dotplot showing variable importance for the top three nodes of all trees in the forest, (3) a static display of one PPtree, (4) a faceted density plot of projected data at each node of the tree, with split point indicated by a vertical line, and (5) a mosaic plot showing the confusion matrix for each node of the tree.  The interaction is driven in three ways, from the boxplot -- when the user selects a point in that display, the corresponding importance variable plot, tree plot from the \verb#PPtreeViz#, tree density displays and mosaic plots are drawn. The tree plot from the \verb#PPtreeViz# is used to visualize the selected tree structure. In addition, the variable importance values are highlighted for each variable for the top three nodes, by default, and the oob error value for the selected tree on the boxplot (both in red).
Additionally it is possible to select specific nodes and update the importance variable plot, density plot and mosaic plot (nodes can be selected on the right corner on this second tab) and finally the interaction can be driven from the variable importance plot.

The selected tree in Figure \ref{tab2} has very small oob error shown as red point in the boxplot. Two variables were use in each node partition, the first four nodes were selected and shown in the importance variable plot, density plot and mosaic plot. In the importance variable plot \verb#height# and \verb#lenght1# were used in the first node partition being \verb#height# the important variable to separate Bream and Parkki (Green and orange in density and mosaic plots) from the rest. In the second node \verb#length1# and \verb#length2# were both evenly important  to separate Bream from Parkki. A similar analysis can be done for other nodes. This tab allow us to understand individual trees in the forest and analyzed which variables are important for the node partition related with the oob error involved in that specific tree.

@@

\begin{figure}[hbpt]
\centering
\includegraphics[width=1\linewidth]{figure/fishT2.pdf}
\caption{Schematic diagram illustrating the interactivity in and between plots for model level exploration panel of the web app. The boxplot and the dotplot of variable importance has click selection, which invokes changes of the corresponding displays. Selecting a point in the boxplot update the node tree information (Shiny Gui) and update all the plots for the three first nodes, also nodes can be selected and update the corresponding plots. Finally selecting a point in the importance variable plot makes a change in the data, which propagates the importance values for other variables in this tree to be highlighted (red), draws the tree, highlights the error value of the tree, and shows the projections and confusion matrix for the three top nodes.}
\label{tab2diag}
\end{figure}
\newpage

\subsection{Performance Comparison}
The third tab (Figure \ref{tab3}) examines the PPF fit and compares the result with a RF fit. There are four displays for each type of model: (1) Variable importance for all trees in the forest (same as in the models tab), (2) a receiver operating characteristic curve (ROC) comparing sensitivity and specificity for each class, (3) oob error by the number of trees to assess complexity, (4) overall variable importance. There is very little interaction on this tab. Users can select to focus on a subset of classes or choose the importance measure to show. Being able to focus on class can help to better understand how well the model performs across classes and the focus can be especially useful for unbalanced data. Examining the oob error by trees enables an assessment of how few trees might be used to provide an equally accurate prediction of future data.

@@

Having better tools to open up black box models will provide a better understanding of the data, the model strengths and weaknesses, and how the model will perform on future data. This visualisation app provides a selection of interactive plots to diagnose PPF models. This shell could be used to make an app for other ensemble classifiers. The philosophy underlying the collection of displays is ``show the model in the data space''  explained in \citet{wickham2015visualizing}. It is not easy to do this, and to completely take this on would require plotting the model in the $p$-dimensional data space. In the simplest approach, as taken here, it means to link the model diagnostics to displays of the data. Then it is possible to probe and query to obtain a better understanding such as finding regions in the data that prove difficult to fit, and detract from the predictive accuracy, or that don't adhere to model assumptions.

@@

The application have some limitations that need to be pointed out. If the number of observation is big, the overploting will be an issue in some of the selected visualizations, and if the number of variables increases, plots like parallel coordinate plot and importance variable plot should include the option to select some variables each time or a different approach must be considered.

There are many possible extensions to the app that could help to become a tool for model refinement: (1) Using the diagnostics to weed out under-performing models in the ensemble; (2) Identifying and boosting models that perform well, particularly if they do well for problematic subsets of the data; (3) Problematic cases could be removed, and ensembles re-fit; (4) Classes as a whole could be aggregated or re-organised as suggested by the model diagnostics, to produce a more effective hierarchical approach to the multiple class problem. Working within the R environment makes all of these desires available using command line outside the app, given the unique IDs of models and cases can be exported from the app.

The app has helped to identify ways to improve the  PPtree algorithm and consequently, the PPF model which especially apply to multiclass problems. Multiple splits for the same class would enable nonlinear classifications. Split criteria tend to place boundaries too close to some groups, due to heteroskedasticity being induced by aggregating classes. Forests are not always better than their constituent trees, and if the trees can be built better, the forest will provide stronger predictions.
